Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-13 paper/quarto_ecml/paper.tex
Add draft conclusions

>>>>

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Although post-hoc
explainability approaches have attracted major research interest, these
cannot guarantee that the explanations agree with the opaque model's
learned representation of data. As a step towards addressing this
challenge, we introduced counterfactual training, a novel training
regime that incentivizes highly-explainable models. Our approach leads
to explanations that are both plausible---compliant with the underlying
data-generating process---and actionable---compliant with user-specified
mutability constraints---and thus meaningful to their recipients.
Through extensive experiments we demonstrate that counterfactual
training satisfies its objectives while preserving the predictive
performance of the trained models. We also find that our approach can be
used to fine-tune conventionally-trained models and achieve similar
gains in explainability. Finally, this work showcases that it is
practical to improve models \emph{and} their explanations at the same
time.
